---
title: "C2. The SLR Model"
output: html_notebook
date: "2024-05-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(
  fig.path = "README_figs/README-"
)
```

## 2.1 Simple linear regression

We are concerned with estimating the population parameters $β_0$ and
$β_1$ of the simple linear regression model 
$$
y = \beta_0 + \beta_1 + \mu
$$ from a random sample of $y$ and $x$. According to Wooldridge (2019,
Section 2.2), the ordinary least squares (OLS) estimators are 
$$
\hat{\beta}_0 = \bar{y} + \hat{\beta}_1x
$$ 
$$
\hat{\beta}_1 = \frac{Cov(x, y)}{Var(x)}
$$ Based on these estimated parameters, the OLS regression line is $$
y = \hat{\beta}_0 + \hat{\beta}_1
$$

### Wooldridge, Example 2.3: CEO Salary and Return on Equity

```{r}
library(wooldridge)
```

```{r message=FALSE}
data(ceosal1, package='wooldridge')
attach(ceosal1)

# manual calculation of OLS coefficients
b1hat <- cov(roe,salary)/var(roe)
b0hat <- mean(salary) - b1hat*mean(roe) 
cat(paste('b0_hat: ', b0hat, '\nb1_hat: ', b1hat))

# "detach" the data frame
detach(ceosal1)
```

Finally, the parameter estimates are calculated. So the OLS regression
line is 
$$
\hat{salary} = 963.1 + 18.5 roe
$$ 
The regression results in a variable CEOregres and then use it as an
argument to abline to add the regression line to the scatter plot

```{r}
data(ceosal1, package='wooldridge')
# OLS regression
CEOregres <- lm( salary ~ roe, data=ceosal1 )
# Scatter plot (restrict y axis limits)
with(ceosal1, plot(roe, salary, ylim=c(0,4000)))
# Add OLS regression line
abline(CEOregres)
```

### Wooldridge, Example 2.4: Wage and Education

```{r}
data(wage1,package='wooldridge')
#OLSregression:
lm(wage~educ,data=wage1)
```

### Wooldridge, Example 2.5: Voting Outcomes and Campaign Expenditures

$$
voteA = \beta_0 +\beta_1 shareA + u
$$

```{r}
data(vote1, package='wooldridge')
vote_reg = lm(voteA ~ shareA, data = vote1)
with(vote1, plot(shareA, voteA))
abline(vote_reg)
```

## 2.2 Coefficients, Fitted Values and Residuals

##### Coefficients

```{r}
vote_reg$coefficients
```

##### Fitted Value $\hat{y_i} = \hat{\beta_0} + \hat{\beta_1}x_1$

##### Residual $\hat{u_i} = y_i - \hat{y_i}$

### Example 2.6: CEO Salary and Return on Equity

```{r}
data(ceosal1, package='wooldridge')
# extract variables as vectors:
sal <- ceosal1$salary
roe <- ceosal1$roe
# regression with vectors:
CEOregres <- lm( sal ~ roe )
# obtain predicted values and residuals
sal.hat <- fitted(CEOregres)
u.hat <- resid(CEOregres)
# Wooldridge, Table 2.2:
cbind(roe, sal, sal.hat, u.hat)[1:15,]
```

### Three properties of OLS statistic

$$
\sum_{i=1}^{n} {u_i} = 0 \rightarrow \bar{\hat{u_i}} = 0
$$ 
$$
\sum_{i=1}^{n} {x_i \hat{u_i}} =0 \rightarrow Cov(x_i,\hat{u_i}) = 0
$$ 
$$
\bar{y} = \hat{\beta_0} + \hat{\beta_1} \cdot \bar{x}
$$
