---
title: "Chapter 2. The Simple Regression ModelModel"
output:
  html_notebook: default
  pdf_document: default
  
editor_options:
  markdown:
    wrap: 72
---

## 2.1 Simple linear regression

We are concerned with estimating the population parameters $β_0$ and
$β_1$ of the simple linear regression model $$
y = \beta_0 + \beta_1 + \mu
$$ from a random sample of $y$ and $x$. According to Wooldridge (2019,
Section 2.2), the ordinary least squares (OLS) estimators are $$
\hat{\beta}_0 = \bar{y} + \hat{\beta}_1x
$$ $$
\hat{\beta}_1 = \frac{Cov(x, y)}{Var(x)}
$$ Based on these estimated parameters, the OLS regression line is $$
y = \hat{\beta}_0 + \hat{\beta}_1
$$

### Wooldridge, Example 2.3: CEO Salary and Return on Equity

```{r}
library(wooldridge)
library(equatiomatic)
library(parameters)
library(gt)
```

```{r message=FALSE}
data(ceosal1, package='wooldridge')
attach(ceosal1)

# manual calculation of OLS coefficients
b1hat <- cov(roe,salary)/var(roe)
b0hat <- mean(salary) - b1hat*mean(roe) 
cat(paste('b0_hat: ', b0hat, '\nb1_hat: ', b1hat))

# "detach" the data frame
detach(ceosal1)
```

Finally, the parameter estimates are calculated. So the OLS regression
line is $$
\hat{salary} = 963.1 + 18.5 roe
$$ The regression results in a variable CEOregres and then use it as an
argument to abline to add the regression line to the scatter plot

```{r}
data(ceosal1, package='wooldridge')
# OLS regression
CEOregres <- lm( salary ~ roe, data=ceosal1 )
# Scatter plot (restrict y axis limits)
with(ceosal1, plot(roe, salary, ylim=c(0,4000)))
# Add OLS regression line
abline(CEOregres)
```

### Wooldridge, Example 2.4: Wage and Education

```{r}
data(wage1,package='wooldridge')
#OLSregression:
lm(wage~educ,data=wage1)
```

### Wooldridge, Example 2.5: Voting Outcomes and Campaign Expenditures

$$
voteA = \beta_0 +\beta_1 shareA + u
$$

```{r}
data(vote1, package='wooldridge')
vote_reg = lm(voteA ~ shareA, data = vote1)
with(vote1, plot(shareA, voteA))
abline(vote_reg)
print_html(model_parameters(vote_reg, summary = TRUE))
```

## 2.2 Coefficients, Fitted Values and Residuals

##### Coefficients

```{r}
vote_reg$coefficients
```

##### Fitted Value $\hat{y_i} = \hat{\beta_0} + \hat{\beta_1}x_1$

##### Residual $\hat{u_i} = y_i - \hat{y_i}$

### Example 2.6: CEO Salary and Return on Equity

```{r}
data(ceosal1, package='wooldridge')
# extract variables as vectors:
sal <- ceosal1$salary
roe <- ceosal1$roe
# regression with vectors:
CEOregres <- lm( sal ~ roe )
# obtain predicted values and residuals
sal.hat <- fitted(CEOregres)
u.hat <- resid(CEOregres)
# Wooldridge, Table 2.2:
cbind(roe, sal, sal.hat, u.hat)[1:15,]
```

### Three properties of OLS statistic

$$
\sum_{i=1}^{n} {u_i} = 0 \rightarrow \bar{\hat{u_i}} = 0
$$ $$
\sum_{i=1}^{n} {x_i \hat{u_i}} =0 \rightarrow Cov(x_i,\hat{u_i}) = 0
$$ $$
\bar{y} = \hat{\beta_0} + \hat{\beta_1} \cdot \bar{x}
$$ \### Wooldridge, Example 2.7: Wage and Education

```{r}
data(wage1, package='wooldridge')
WAGEregres <- lm(wage ~ educ, data=wage1)
# obtain coefficients, predicted values and residuals
b.hat <- coef(WAGEregres)
wage.hat <- fitted(WAGEregres)
u.hat <- resid(WAGEregres)
# Confirm property (1):
mean(u.hat)
# Confirm property (2):
cor(wage1$educ , u.hat)
# Confirm property (3):
mean(wage1$wage)
b.hat[1] + b.hat[2] * mean(wage1$educ)
```

## 2.3 Goodness of fit

The total sum of squares (SST), explained sum of squares (SSE) and
residual sum of squares (SSR) can be written as

$$
SST = \sum_{i=1}^{n}{(y_i - \bar{y})^2} = (n-1) \cdot Var(y)
$$

$$
SSE = \sum_{i=1}^{n}{(\hat{y_i} - \bar{y})^2} = (n-1) \cdot Var(\hat{y})
$$ $$
SSE = \sum_{i=1}^{n}{\hat{u_i}^2} = (n-1) \cdot Var(\hat{u})
$$ Because (n − 1) cancels out, it can be equivalently written as $$
R^2 = \frac{Var(\hat{y})}{Var(y)} = 1 -\frac{Var(\hat{u})}{Var(y)}
$$

### Example 2.8: CEO Salary and Return on Equity

```{r}
data(ceosal1, package='wooldridge')
CEOregres <- lm( salary ~ roe, data=ceosal1 )
# Calculate predicted values & residuals:
sal.hat <- fitted(CEOregres)
u.hat <- resid(CEOregres)
# Calculate R^2 in three different ways:
sal <- ceosal1$salary
var(sal.hat) / var(sal)

1 - var(u.hat) / var(sal)

cor(sal, sal.hat)^2
```

### Example 2.9: Voting Outcomes and Campaign Expenditures

```{r}
data(vote1, package='wooldridge')
VOTEres <- lm(voteA ~ shareA, data=vote1)
# Summary of the regression results
summary(VOTEres)
```

## 2.4. Nonlinearities

### Wooldridge, Example 2.10: Wage and Education

```{r}
data(wage1, package='wooldridge')
# Estimate log-level model
lm( log(wage) ~ educ, data=wage1 )
```

### Example 2.11: CEO Salary and Firm Sale

```{r}
data(ceosal1, package='wooldridge')
# Estimate log-log model
lm( log(salary) ~ log(sales), data=ceosal1 )
```

## 2.5. Regression through the Origin and Regression on a Constant

###### Through the Origin

lm(y \~ 0 + x)

###### Through a Constant

lm(y \~ 1)

##### Example Code

```{r}
data(ceosal1, package='wooldridge')
# Usual OLS regression:
(reg1 <- lm( salary ~ roe, data=ceosal1))
# Regression without intercept (through origin):
(reg2 <- lm( salary ~ 0 + roe, data=ceosal1))
# Regression without slope (on a constant):
(reg3 <- lm( salary ~ 1 , data=ceosal1))
# average y:
mean(ceosal1$salary)
# Scatter Plot with all 3 regression lines
plot(ceosal1$roe, ceosal1$salary, ylim=c(0,4000))
abline(reg1, lwd=2, lty=1)
abline(reg2, lwd=2, lty=2)
abline(reg3, lwd=2, lty=3)
legend("topleft",c("full","through origin","const only"),lwd=2,lty=1:3)

```

## 2.6. Expected Values, Variances, and Standard Errors

-   SLR.1: Linear population regression function: $y = β_0 + β_1x + u$
-   SLR.2: Random sampling of x and y from the population
-   SLR.3: Variation in the sample values $x1, ..., x_n$
-   SLR.4: Zero conditional mean: $E(u|x) = 0$
-   SLR.5: Homoscedasticity: $Var(u|x) = σ$
